---
title: "scplainer: reanalysis of the plexDIA dataset (Derks et al. 2022)"
author:
    - Christophe Vanderaa, Computational Biology, UCLouvain
    - Laurent Gatto, Computational Biology, UCLouvain
output:
    BiocStyle::html_document:
        self_contained: yes
        toc: true
        toc_float: true
        toc_depth: 2
        code_folding: show
date: "`r BiocStyle::doc_date()`"
vignette: >
    %\VignetteIndexEntry{scplainer: derks2022 data analysis}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
## Options for Rmarkdown compilation
knitr::opts_chunk$set(
    fig.width = 7,
    fig.height = 5,
    fig.align = "center",
    message = FALSE,
    warning = FALSE,
    collapse = TRUE,
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
## Time the compilation
timeStart <- Sys.time()
```

# Introduction

In this vignette, we will analyse the `derks2022` dataset using the 
*scplainer* approach. The
data were acquired using the plexDIA protocol 
([Derks et al. 2022](http://dx.doi.org/10.1038/s41587-022-01389-w)).
The authors performed mTRAQ-based labelling in DIA mode, and showed
that plexDIA could increase the cell throughput while benefitting from
the advantages of DIA.

Before delving into the scplainer workflow, we recommend first 
familiarising with the `scp` and `QFeatures` packages (see 
[tutorials](https://uclouvain-cbio.github.io/scp/index.html)). 

# Packages and data

We rely on several packages to compile this vignette.

```{r, warning=FALSE, message=FALSE}
## Core packages
library("scp")
library("scpdata")
## Utility packages
library("ggplot2")
library("patchwork")
library("dplyr")
library("camprotR")
```

The data set is available from the `scpdata` package. 

```{r}
derks <- derks2022()
```

The data set contains 186 samples, including 12 negative control 
samples, 12 bulk samples, 52 monocytes (U-937), 55 melanoma cells, and
55 pancreatic ductal adenocarcinoma cells (PDAC). The samples were 
labelled with mTRAQ-3 acquired either with a Q-Exactive (n = 153) or a
timsTOF-SCP instrument (n = 33). 

# Minimal data processing

The minimal data processing workflow in *scplainer* consists of 5 main steps:

- Data cleaning
- Feature quality control
- Sample quality control
- Peptide data assembly
- Log2-transformation

## Cleaning data

The data available in `scpdata` were provided by the authors and were 
analysed with DIA-NN. We will start with the precursor data and take 
advantage of DIA-NN's capability to transfer precursor identities 
across mTRAQ labels within the same runs. These data are available in 
the `qe_prec_extracted` and `tims_prec_extracted` assays. 

```{r}
derks <- derks[, , c("qe_prec_extracted", "tims_prec_extracted")]
```

Next, we clean the cell type annotation and remove the `_t` tags 
contained in cell type annotations that refer to samples acquired with
timsTOF-SCP. Note that samples annotated with a `_DB` are bulk 
samples. 

```{r}
derks$Celltype <- sub("_t", "", derks$Celltype)
table(derks$Celltype)
```

The instrument information is available in a dedicated descriptor.

```{r}
table(derks$Instrument)
```

The data were acquired using mTRAQ-3 labelling ($\delta$0, $\delta$4 
and $\delta$8).

```{r}
table(derks$Label)
```

The data were acquired as part of 59 MS acquisition batches. 

```{r}
length(unique(derks$Set))
```

Next, we remove feature annotations that won't be used in the remainder of
the vignette. This is to avoid overcrowding of the annotation tables
later in the vignette.

```{r}
requiredRowData <- c(
    "Protein.Group", "Protein.Ids", "Protein.Names", "Genes", 
    "First.Protein.Description", "Proteotypic", "Stripped.Sequence",
    "Modified.Sequence", "Precursor.Charge", "Precursor.Id"
)
derks <- selectRowData(derks, requiredRowData)
```

We replace zeros with missing values. A zero may be a true zero (the feature
is not present in the sample) or because of technical limitations (due
to the technology or the computational pre-processing). Because we are
not able to distinguish between the two, zeros should be replaced with
`NA`.

```{r}
derks <- zeroIsNA(derks, i = names(derks))
```

Below is an overview of the cleaned `Qfeatures` object used as input 
data for the data analysis. 

```{r}
derks
```

## Feature quality control

We remove contaminants peptides that may propagate technical
artefacts and bias data modelling. To identify contaminants we 
retrieve the UniProt sequences in cRAP through the `camprotR` package.
We also remove any peptide that maps to a keratin sequence. 

```{r}
contaminants <- paste(get_ccp_crap(), collapse = "|")
keratins <- "KRT|^K\\d.\\d"
derks <- filterFeatures(derks, ~ !grepl(contaminants, Protein.Ids) &
                          !grepl(keratins, Protein.Names))
```

## Sample quality control 

We next remove low-quality cells. The quality control criteria are:

- We remove samples with low number of detected peptides. The 
criterion is computed as follows:

```{r}
derks <- countUniqueFeatures(
    derks, i = names(derks),
    groupBy = "Stripped.Sequence", colDataName = "NumberPeptides"
)
```

- We remove samples with low median intensity. The metric (note we 
will later use it for normalisation) is computed as follows:

```{r}
MedianIntensity <- lapply(experiments(derks), function(x) {
    out <- colMedians(log(assay(x)), na.rm = TRUE)
    names(out) <- colnames(x)
    out
})
names(MedianIntensity) <- NULL
MedianIntensity <- unlist(MedianIntensity)
colData(derks)[names(MedianIntensity), "MedianIntensity"] <- MedianIntensity
```

- We remove the samples that have a high median coefficient of 
variation (CV). The CV is computed within each sample, by grouping
the peptides that belong to the same protein or protein group. 
Computing the median CV requires peptide normalisation (`norm = "SCoPE2"`)
to avoid that differences in ionisation obscure the quantification accuracy. 
This is computed as follows:

```{r}
derks <- medianCVperCell(
    derks, i = names(derks),
    groupBy = "Protein.Ids", nobs = 3,
    norm = "SCoPE2",
    colDataName = "MedianCV"
)
```

We plot the metrics used to perform sample quality control. Since two
different instruments were used, we expect different metric 
distributions for each instrument and we plot the metrics separately.

```{r qc_plot, fig.width=8, fig.height=4}
pl <- list()
for (i in unique(derks$Instrument)) {
    pl[[i]] <- data.frame(colData(derks)) |>
        dplyr::filter(Instrument == i) |>
        ggplot() +
        aes(y = MedianIntensity,
            x = NumberPeptides,
            color = MedianCV,
            shape = Celltype) +
        ggtitle(i) +
        geom_point(size = 2) +
        scale_color_continuous(type = "viridis")
}
wrap_plots(pl)
```

There are a few suspicious cells with a low number of detected 
peptides, a low median intensity and a high median CV. For the 
Q-Exactive dataset, we remove the group of cells that cluster with the
negative control samples, that are cells with less than 750 peptides
and a median intensity lower than 9.4. For the timsTOF-SCP dataset, we
remove the group cells characterised by a median intensity lower than
6.5. We apply the QC filter but also remove any sample that is not a
single cell.

```{r}
derks$passQC <- (derks$Instrument == "Q-Exactive" &
                    derks$MedianIntensity > 9.4 &
                    derks$NumberPeptides > 750) |
                (derks$Instrument == "timsTOFSCP" &
                    derks$MedianIntensity > 6.5)
derks$isSingleCell <- !grepl("Neg|DB$", derks$Celltype)
derks <- subsetByColData(derks, derks$passQC & derks$isSingleCell)
```

## Peptide data assembly

For now, we worked with precursor-level data. We here aggregate 
precursors into peptides. However, we need to solve a minor issue in 
the feature annotations. When aggregating peptides, we will only keep
the feature annotations that have the common data. However, the 
protein groups differ among a few precursor from the same peptide.
Hence, the protein groups and the associated gene names are not 
constant for some peptides that are removed during aggregation. To 
avoid this, we ensure consistant protein groups across precursor from 
the same peptide using majority vote.

```{r}
ppMap <- rbindRowData(derks, i = names(derks)) %>%
    data.frame %>%
    group_by(Stripped.Sequence) %>%
    ## The majority vote happens here
    summarise(
        Protein.Group = names(sort(table(Protein.Group), decreasing = TRUE))[1],
        Genes = names(sort(table(Genes), decreasing = TRUE))[1]
    ) %>%
    dplyr::select(Stripped.Sequence, Protein.Group, Genes)
consensus <- lapply(names(derks), function(i) {
    ind <- match(rowData(derks)[[i]]$Stripped.Sequence, ppMap$Stripped.Sequence)
    DataFrame(
        Protein.Group = ppMap$Protein.Group[ind],
        Genes = ppMap$Genes[ind]
    )
})
names(consensus) <- names(derks)
rowData(derks) <- consensus
```

We then can safely perform peptide aggregation.

```{r feature_aggregation, message=FALSE}
peptideAssays <- paste0("peptides_", names(derks))
derks <- aggregateFeatures(
    derks, i = names(derks),
    fcol = "Stripped.Sequence",
    name = peptideAssays,
    fun = colMedians, na.rm = TRUE
)
```

Moreover, the data are stored in separate assays for each instrument.
We here combine the two assays into one.

```{r}
derks <- joinAssays(derks, i = peptideAssays, name = "peptides")
```

## Log-transformation 

We log2-transform the quantification data. 

```{r}
derks <- logTransform(derks, i = "peptides", name = "peptides_log")
```

Here is an overview of the data processing:

```{r}
plot(derks)
```

# Data modelling

Model the data using scplainer, a linear regression model implemented in `scp`.
The model is applied on a `SingleCellExperiment` so we extract it from
the processed data set along with the `colData`

```{r}
sce <- getWithColData(derks, "peptides_log")
```

First, we must specify which variables to include in the model. We 
here include 4 variables:

- `MedianIntensity`: this is the normalisation factor used to correct 
  for cell-specific technical differences.
- `Set`: this is used to perform batch correction. We consider each 
  acquisition run to be a batch. 
- `Label`: this is used to correct for mTRAQ effects.
- `Celltype`: this is the biological variable of interest. It
  captures the difference between macrophages and monocytes

`scpModelWorkflow()` fits linear regression models to the data, where
the model is adapted for each peptide depending on its pattern of 
missing values.

```{r data_modelling, results='hide'}
sce <- scpModelWorkflow(
    sce,
    formula = ~ 1 + ## intercept
        ## normalisation
        MedianIntensity +
        ## batch effects
        Set + Label +
        ## biological variability
        Celltype)
```

Once the model is prepared, we can explore the distribution of the n/p
ratios.

```{r}
scpModelFilterPlot(sce)
```

About 40 % of the peptides do not have sufficient observations to estimate the
model. You could consider $n/p$ a rough average of the number of
replicates per parameter to fit (for categorical variables, the number
of replicates per group). We recommend moving the threshold away from
1 to increase statistical power and remove noisy peptides. This comes
of course at the cost of less peptides included in the analysis.

# Model analysis

The model analysis consists of three steps: 

- Analysis of variance
- Differential abundance analysis
- Component analysis

## Analysis of variance

The analysis of variance explores the proportion of data captures by each
variable in the model. 

```{r variance_analysis}
(vaRes <- scpVarianceAnalysis(sce))
vaRes[[1]]
```

The results are a list of tables, one table for each variable. Each 
table reports for each peptide the variance captures (`SS`), the 
residual degrees of freedom for estimating the variance (`df`) and the
percentage of total variance explained (`percentExplainedVar`). To
better explore the results, we add the annotations available in the
`rowData`. 

```{r}
vaRes <- scpAnnotateResults(
    vaRes, rowData(sce), by = "feature", by2 = "Stripped.Sequence"
)
```

By default, we explore the variance for all peptides combined:

```{r, fig.width=2.5, fig.height=4, out.width = "25%"}
scpVariancePlot(vaRes)
```

We explore the top 20 peptides that are have the highest percentage of
variance explained by the biological variable (top) or by the batch
variable (bottom).

```{r, fig.width=10, fig.height=8}
scpVariancePlot(
    vaRes, top = 20, by = "percentExplainedVar", effect = "Celltype",
    decreasing = TRUE, combined = FALSE
) +
    scpVariancePlot(
    vaRes, top = 20, by = "percentExplainedVar", effect = "Set",
    decreasing = TRUE, combined = FALSE
) +
    plot_layout(ncol = 1, guides = "collect")
```

We can also group these peptide according to their protein. We simply
need to provide the `fcol` argument.

```{r, fig.width=10, fig.height=8}
scpVariancePlot(
    vaRes, top = 20, by = "percentExplainedVar", effect = "Celltype",
    decreasing = TRUE, combined = FALSE, fcol = "Genes"
) +
    scpVariancePlot(
    vaRes, top = 20, by = "percentExplainedVar", effect = "Set",
    decreasing = TRUE, combined = FALSE, fcol = "Genes"
) +
    plot_layout(ncol = 1, guides = "collect")
```

## Differential abundance analysis

Next, we explore the model output to understand the differences 
between the cell types. The difference of interest is
specified using the `contrast` argument. The first element points to 
the variable to test and the two following element are the groups of
interest to compare. You can provide multiple contrast in a list.

```{r}
(daRes <- scpDifferentialAnalysis(
    sce, contrast = list(
        c("Celltype", "Melanoma", "PDAC"),
        c("Celltype", "Melanoma", "U-937"),
        c("Celltype", "PDAC", "U-937")
    )
))
daRes[[1]]
```

Similarly to analysis of variance, the results are a list of tables, one 
table for each contrast. Each table reports for each peptide the 
estimated difference between the two groups, the standard error 
associated to the estimation, the degrees of freedom, the 
t-statistics, the associated p-value and the p-value FDR-adjusted for
multiple testing across all peptides. Again, to better explore the 
results, we add the annotations available in the `rowData`. 

```{r}
daRes <- scpAnnotateResults(
    daRes, rowData(sce), 
    by = "feature", by2 = "Stripped.Sequence"
)
```

We then visualize the results using a volcano plot. The function below
return a volcano plot for each contrast. 

```{r, fig.width=10, fig.height=4.5, warning=FALSE}
wrap_plots(scpVolcanoPlot(daRes))
```

To help interpretation of the results, we will label the peptides with
their protein name. Also we increase the number of labels shown on the plot.
Finally, we can add colors to the plot. For instance, let's explore 
the impact of the number of observations using the $n/p$ ratio. We
create a new annotation table, add it to the results and redraw the 
plot. The $n/p$ ratio is retrieved using scpModelFilterNPRatio

```{r, fig.width=10, fig.height=4.5}
np <- scpModelFilterNPRatio(sce)
daRes <- scpAnnotateResults(
    daRes, data.frame(feature = names(np), npRatio = np),
    by = "feature"
)
wrap_plots(scpVolcanoPlot(
    daRes, top = 30, textBy = "Genes", 
    pointParams = list(aes(colour = npRatio))
))
```

As expected, higher number of observations (higher $n/p$) lead to
increased statistical power and hence to more significant results.

Finally, we offer functionality to report results at the protein 
level. 

```{r, fig.width=10, fig.height=4.5, warning=FALSE}
scpDifferentialAggregate(daRes, fcol = "Genes") |>
    scpVolcanoPlot(top = 30, textBy = "Genes") |>
    wrap_plots()
```

## Component analysis

Finally, we perform component analysis to link the modelled effects to
the cellular heterogeneity. We here run an APCA+ (extended 
ANOVA-simultaneous principal component analysis) for the sample type 
effect. In other words, we perform a PCA on the data that is capture
by the sample type variable along with the residuals (unmodelled 
data).

```{r component_analysis}
(caRes <- scpComponentAnalysis(
    sce, ncomp = 2, method = "APCA", effect = "Celltype"
))
```

The results are contained in a list with 2 elements. `bySample`
contains the PC scores, that is the component results in sample space.
`byFeature` contains the eigenvectors, that is the component results
in feature space. 

```{r}
caRes$bySample
```

Each of the two elements contains components results
for the data before modelling (`unmodelled`), for the residuals or for
the APCA on the sample type variable (`APCA_Celltype`). 

```{r}
caRes$bySample[[1]]
```

Each elements is a table with the computed componoents. Let's explore
the component analysis results in cell space. Similarly to the
previous explorations, we annotate the results.

```{r}
caResCells <- caRes$bySample
sce$cell <- colnames(sce)
caResCells <- scpAnnotateResults(caResCells, colData(sce), by = "cell")
```

We then generate the component plot, colouring by `Celltype`. To 
assess the impact of batch effects, we shape the points according to 
the plate batch (cf intro) as well.

```{r, fig.width=9, fig.height=3}
scpComponentPlot(
    caResCells,
    pointParams = list(aes(colour = Celltype, shape = Instrument))
) |>
    wrap_plots() +
    plot_layout(guides = "collect")
```

While the data before modelling is mainly driven by batch effects
(instruments are separated on the first PC), the APCA clearly 
separates the three cell populations. The PCA on the residuals 
suggests that no additional unmodelled effects influence the data
globally, indicating that our model accurately captures the variation
in the data.

We use the same approach to explore the component results in feature
space. 

```{r}
caResPeps <- caRes$byFeature
caResPeps <- scpAnnotateResults(
    caResPeps, rowData(sce), by = "feature", by2 = "Stripped.Sequence"
)
```

We plot the compenents in peptide-space. 

```{r, fig.width=9, fig.height=3}
plCApeps <- scpComponentPlot(
    caResPeps, pointParams = list(size = 0.8, alpha = 0.4)
)
wrap_plots(plCApeps)
```

We can also combine the exploration of the components in cell and
peptide space. This is possible thanks to biplots.

```{r, fig.width=9, fig.height=3}
biplots <- scpComponentBiplot(
    caResCells, caResPeps, 
    pointParams = list(aes(colour = Celltype, shape = Instrument)),
    labelParams = list(size = 1.5, max.overlaps = 20),
    textBy = "Genes", top = 20
)
wrap_plots(biplots, guides = "collect")
```

Finally, we offer functionality to aggregate the results at the 
protein level instead of the peptide level. 

```{r, fig.width=9, fig.height=3}
caResProts <- scpComponentAggregate(caResPeps, fcol = "Genes")
biplots <- scpComponentBiplot(
    caResCells, caResProts, 
    pointParams = list(aes(colour = Celltype, shape = Instrument)),
    labelParams = list(size = 1.5, max.overlaps = 20),
    textBy = "Genes", top = 20
)
wrap_plots(biplots, guides = "collect")
```

# Interactive visualisation with iSEE

You can manually explore the data through an interactive interface 
thanks to using `iSEE`: 

```{r, eval=FALSE}
library("iSEE")
iSEE(sce)
```

# Session information {-}

```{r setup2, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "",
    crop = NULL
)
```

```{r sessioninfo, echo=FALSE}
sessionInfo()
```

# Citation

```{r citation}
citation("scp")
```

# License

This vignette is distributed under a 
[CC BY-SA license](https://creativecommons.org/licenses/by-sa/2.0/) 
license.
